BTIC.dat Differential Methylation Analysis
========================================================
## By: Sumaiya Islam
## Date: August 30, 2016

### Script contents:
- Differential methylation analysis of BTIC.dat_Combat 

### A. Set up working directory & packages


We will initially set our working directory and load our libraries.
```{r}
setwd("~/BTIC_GBM")
library(methylumi)
library(gplots)
library(marray)
library(lumi)
library(lattice)
library("RColorBrewer")
library(knitr)
library(xtable)
library(ggplot2)
library(reshape)
library(mclust)
library(limma)
library(reshape)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(gridExtra)
library(sva)
library(rama)
```

### B. Load filtered dataset. 

We will load the ComBat-corrected cancer.dat dataset and then filter out the technical replicates to yield a final dataset comprised of 37 pairs of matched BTIC and tumour methylation profiles. 

```{r}
# load ComBat-corrected data
# load(file="BTIC.dat_Combat.RData") # Combat corrected data
# dim(BTIC.dat_Combat) # probes = 420195, samples = 44
# 
# # filter out all technical replicates
# reps<-c("67_cell_rep", "147_cell_rep", "73_cell_rep", "12_cell_rep")
# BTIC.dat.cor<- BTIC.dat_Combat[,!(sampleNames(BTIC.dat_Combat)) %in% reps]
# dim(BTIC.dat.cor) # probes = 420,195, samples = 40
# sampleNames(BTIC.dat.cor)
# 
# # add subtype to meta data (according to Yaoqing, all BTIC samples are classical subtype except for 47_cell, 84_cell, 92_cell, and 142_cell that show proneural subtype). 
# meta.BTIC<-pData(BTIC.dat.cor)
# rownames(meta.BTIC) # check which cell lines are in final dataset
# meta.BTIC$Subtype<-c(rep("classical", times = 15), "proneural", rep("classical", times = 24))
# meta.BTIC$Subtype<-as.factor(meta.BTIC$Subtype)
# pData(BTIC.dat.cor)<-meta.BTIC

# save final corrected BTIC dataset
# save(BTIC.dat.cor, file = "BTIC.dat.cor.RData")
load("BTIC.dat.cor.RData")
revised.meta<-read.csv(file="Meta_Data_BTIC.dat_Jan_2017.csv", header = TRUE, row.names = 1)
revised.meta$Sample<-rownames(revised.meta)
identical(colnames(exprs(BTIC.dat.cor)), revised.meta$Sample) # TRUE
pData(BTIC.dat.cor)<-revised.meta
dim(BTIC.dat.cor)
```

### C. Differential Methylation Analysis

#### PCA: Heat scree plot function
```{r}
### Function of association meta variable with PC (ANOVA)
heat_scree_plot<-function(Loadings, Importance, Num, Order){
  adjust<-1-Importance[1]
  pca_adjusted<-Importance[2:length(Importance)]/adjust
  pca_df<-data.frame(adjusted_variance=pca_adjusted, PC=seq(1:length(pca_adjusted)))
  
  scree<-ggplot(pca_df[which(pca_df$PC<Num),],aes(PC,adjusted_variance))+geom_bar(stat = "identity",color="black",fill="grey")+theme_bw()+
        theme(axis.text = element_text(size =12),
              axis.title = element_text(size =15),
              plot.margin=unit(c(1,1.5,0.2,2.25),"cm"))+ylab("Variance")+
    scale_x_continuous(breaks = seq(1,Num,1))
  
  #### Heat
  ## correlate meta with PCS
  ## Run anova of each PC on each meta data variable

  aov_PC_meta<-lapply(1:ncol(meta_categorical), function(covar) sapply(1:ncol(Loadings), function(PC) summary(aov(Loadings[,PC]~meta_categorical[,covar]))[[1]]$"Pr(>F)"[1]))
  cor_PC_meta<-lapply(1:ncol(meta_continuous), function(covar) sapply(1:ncol(Loadings), function(PC) (cor.test(Loadings[,PC],as.numeric(meta_continuous[,covar]),alternative = "two.sided", method="spearman", na.action=na.omit)$p.value)))
  names(aov_PC_meta)<-colnames(meta_categorical)
 names(cor_PC_meta)<-colnames(meta_continuous)
  aov_PC_meta<-do.call(rbind, aov_PC_meta)
 cor_PC_meta<-do.call(rbind, cor_PC_meta)
 aov_PC_meta<-rbind(aov_PC_meta, cor_PC_meta)
  aov_PC_meta<-as.data.frame(aov_PC_meta)
  #adjust
  aov_PC_meta_adjust<-aov_PC_meta[,2:ncol(aov_PC_meta)]
  
    
  #reshape
  avo<-aov_PC_meta_adjust[,1:(Num-1)]
  avo_heat_num<-apply(avo,2, as.numeric)
  avo_heat<-as.data.frame(avo_heat_num)
  colnames(avo_heat)<-sapply(1:(Num-1), function(x) paste(x, sep=""))
  avo_heat$meta<-rownames(avo)
  avo_heat_melt<-melt(avo_heat, id=c("meta"))
  
  # cluster meta data
  ord <- Order
  meta_var_order<-unique(avo_heat_melt$meta)[rev(ord)]
  avo_heat_melt$meta <- factor(avo_heat_melt$meta, levels = meta_var_order)
  
  # color if sig
  avo_heat_melt$Pvalue<-sapply(1:nrow(avo_heat_melt), function(x) if(avo_heat_melt$value[x]>=0.9){">=0.9"}else{
   if(avo_heat_melt$value[x]>=0.5){">=0.5"}else{
     if(avo_heat_melt$value[x]>=0.1){">=0.1"}else{"<0.1"}}})
  avo_heat_melt$Pvalue<-sapply(1:nrow(avo_heat_melt), function(x) if(avo_heat_melt$value[x]<=0.001){"<=0.001"}else{
     if(avo_heat_melt$value[x]<=0.01){"<=0.01"}else{
       if(avo_heat_melt$value[x]<=0.05){"<=0.05"}else{">0.05"}}})
  
  heat<-ggplot(avo_heat_melt, aes(variable,meta, fill = Pvalue)) +
  geom_tile(color = "black",size=0.5) +
  theme_gray(8)+scale_fill_manual(values=c("#084594","#4292c6","#9ecae1","#deebf7"))+
      theme(axis.text = element_text(size =10, color="black"),
            axis.text.x = element_text(),
          axis.title = element_text(size =15),
          legend.text = element_text(size =14),
          legend.title = element_text(size =12),
          legend.position = c(1, 0), legend.justification = c(1,0),
          plot.margin=unit(c(0,2.25,1,1),"cm"))+
    xlab("Principal Component")+ylab(NULL)
  
  grid.arrange(scree, heat, ncol=1)
}

# run PCA
betas.BTIC<-betas(BTIC.dat.cor)

PCA_full<-princomp(betas.BTIC[complete.cases(betas.BTIC),])
Loadings<-as.data.frame(unclass(PCA_full$loadings))
vars <- PCA_full$sdev^2
Importance<-vars/sum(vars)
adjust<-1-Importance[1]
pca_adjusted<-Importance[2:length(Importance)]/adjust
(pca_df<-data.frame(adjusted_variance=pca_adjusted, PC=seq(1:length(pca_adjusted))))

#Specify which covariates are categorical and/or continuous
meta<-pData(BTIC.dat.cor)
meta_categorical<-meta[,c("Diagnosis", "Parental_Tumour_Subtype", "Subtype", "Batch", "Sentrix_ID", "Row", "Sex", "Subtype")]  # input column numbers in meta that contain categorical variables
meta_continuous<- meta[,c("Age")]
meta_continuous<-data.frame(meta_continuous)
colnames(meta_continuous)<-c("Age")
colnames(meta_categorical)<-c("Diagnosis","Parental Tumour Subtype", "Subtype", "Batch", "Chip", "Row", "Sex", "BTIC Subtype")
# Specify the number of PCs you want shown (usually # of samples in the dataset)
Num<-20
# Designate what order you want the variables to appear (continuous variables rbinded to categorical variables in function)
Order<-1:9

#Apply function on PCA results, pulls in the meta data and beta values from above and represent as heat scree plot
heat_scree_plot(Loadings, Importance, Num, Order)

### in ggplot need to adjust for probe offset PC so plot Comp.2 vs Comp.3 to represent PC1 vs PC2
ggplot(Loadings,aes(Comp.2, Comp.3)) + 
  geom_point(aes(fill = factor(BTIC.dat.cor$Diagnosis)), colour="black", shape=21, size=5) + 
  scale_fill_manual(values=c("forestgreen",  "blue"), name="Diagnosis") +
  theme_bw() +
  xlab("PC1 (10.2% of the variance)") + 
  ylab("PC2 (7.6% of the variance)") 
```

To assess parental tumour subtype classification in BTIC DNA methylation profiles, we will subset down to only samples in which parental tumour subtype is known (since many remain unassigned)

```{r}
# specify missing values and subset down to just 
meta<-pData(BTIC.dat.cor)
meta$Parental_Tumour_Subtype<-as.character(meta$Parental_Tumour_Subtype)
meta[meta=='NaN'] <- NA
is.na(meta[,"Parental_Tumour_Subtype"])
meta.subtype<-meta[complete.cases(meta$Parental_Tumour_Subtype),]
complete_samples<-as.character(meta.subtype$Sample)

# subset DNA methylation data down to complete_samples
head(betas.BTIC.subset<-betas.BTIC[,complete_samples])
identical(colnames(betas.BTIC.subset), rownames(meta.subtype)) # TRUE

# run PCA on subsetted data
PCA_full<-princomp(betas.BTIC.subset[complete.cases(betas.BTIC.subset),])
Loadings<-as.data.frame(unclass(PCA_full$loadings))
vars <- PCA_full$sdev^2
Importance<-vars/sum(vars)
adjust<-1-Importance[1]
pca_adjusted<-Importance[2:length(Importance)]/adjust
(pca_df<-data.frame(adjusted_variance=pca_adjusted, PC=seq(1:length(pca_adjusted))))

#Specify which covariates are categorical and/or continuous
meta.subtype$Parental_Tumour_Subtype<-as.factor(meta.subtype$Parental_Tumour_Subtype)
meta_categorical<-meta[,c("Diagnosis", "Batch", "Sentrix_ID", "Row", "Sex", "Subtype", "Parental_Tumour_Subtype")]  # input column numbers in meta that contain categorical variables
meta_continuous<- meta[,c("Age")]
meta_continuous<-data.frame(meta_continuous)
colnames(meta_continuous)<-c("Age")
colnames(meta_categorical)<-c("Diagnosis","Batch", "Chip", "Row", "Sex", "BTIC Subtype", "Parental Tumour Subtype")
# Specify the number of PCs you want shown (usually # of samples in the dataset)
Num<-32
# Designate what order you want the variables to appear (continuous variables rbinded to categorical variables in function)
Order<-1:8

#Apply function on PCA results, pulls in the meta data and beta values from above and represent as heat scree plot
heat_scree_plot(Loadings, Importance, Num, Order)
```

#### Beta Distributions
```{r}
# Beta_Data<-betas(BTIC.dat.cor)
# dim(Beta_Data)
# Beta_sample<-Beta_Data[sample(1:418740, 10000),] # take a random sample of 10,000 probes to plot distribution, otherwise this plot takes forever for dataset with all probes! 
# ## Beta distribtuions 
# Beta_sample_melted<- melt(Beta_sample)
# #remove NAs before plotting (otherwise get many non-inifnite warnings)
# Beta_Plot<-Beta_sample_melted[which(!(is.na(Beta_sample_melted$value))),]
# head(Beta_Plot)
# #add meta
# meta<-pData(BTIC.dat.cor)
# meta$Sample_name<-rownames(meta)
# Beta_Plot<-merge(Beta_Plot,meta, by.x="X2", by.y="Sample_name")
# head(Beta_Plot)
# ggplot(Beta_Plot, aes(value, group=X2, color=Sample_Group))+
#                    geom_density()+theme_bw()
```

#### Hierarchical Clustering
```{r fig.width=10, fig.height=3}
# Beta_Data<-betas(BTIC.dat.cor)
# # remove rows with NAs
# beta_cluster<-Beta_Data[complete.cases(Beta_Data),]
# identical(colnames(Beta_Data), rownames(meta.BTIC))
# # Emulate ggplot default colours from beta distributions
# gg_color_hue <- function(n) {
#   hues = seq(15, 375, length=n+1)
#   hcl(h=hues, l=65, c=100)[1:n]
# }
# 
# # plot clustering with color function
#     plotHclustColors <- function(matrix,leafcolor) {
#       colnames(matrix) <- leafcolor
#       d <- dist(t(matrix))
#       hc <- hclust(d, method = "average") #single, complete, average, ward
#      # color<-rep(brewer.pal(8,"Dark2"),ceiling(length(unique(leafcolor))/12))
#      color<-c("#00B0F6", "#F8766D")
#       labelColors <- color[sample(1:length(color),length(unique(leafcolor)))]
#       colLab <- function(n) {
#         if (is.leaf(n)) {
#           a <- attributes(n)
#           labCol <- labelColors[which(unique(leafcolor) == a$label)]
#           attr(n, "nodePar") <- c(a$nodePar, lab.col=labCol)
#         }
#         n
#       }
#       clusDendro <- dendrapply(as.dendrogram(hc), colLab)
#       plot(clusDendro)
#     }
# 
# # Plot dendrogram
# par(cex=0.7)
# meta$Sample_ID<-rownames(meta)
# plotHclustColors(beta_cluster, meta.BTIC$Patient_ID) #all samples


## Heatmap
# define colour palette
BuPu <- colorRampPalette(brewer.pal(10,"RdYlBu")[1:10])(30)
cor <- cor(betas(BTIC.dat.cor), use = "pairwise.complete.obs")
# define coloured variables
str(pData(BTIC.dat.cor))
colnames(pData(BTIC.dat.cor))
DiagnosisCol <- as.character (as.numeric(factor(BTIC.dat.cor$Diagnosis)))
DiagnosisCol <- gsub("1", "forestgreen",  gsub("2", "blue", DiagnosisCol))
PSubtypeCol<- as.numeric(factor(BTIC.dat.cor$Parental_Tumour_Subtype))
PSubtypeCol <- gsub("1", "#F8766D",  gsub("2", "#7CAE00", gsub("3", "#00BFC4", gsub("4", "#C77CFF", PSubtypeCol))))
heatmap.2(cor,
          trace = "none", col = BuPu, dendrogram = "both",
          cexRow = 1, margins=c(10,10),RowSideColors = PSubtypeCol,
          ColSideColors = DiagnosisCol, cexCol = 1, key=TRUE, keysize = 1.5)
legend("topright",legend = c("gbm", "gbm_r"), 
    col = c("forestgreen", "blue"), 
    lty= 1,             
    lwd = 8            
)

legend("bottomleft",legend = c("Classical", "Mesenchymal", "Proneural", "Unknown"), 
    col = c("#F8766D", "#7CAE00", "#00BFC4", "#C77CFF"), 
    lty= 1,             
    lwd = 8            
)
```


### Try hierarchial clustering on only variable probes?
