Cancer.Dat Epigenetic Age Analysis
================================================================
## By: Sumaiya Islam
## Date: Oct 1, 2017

### Script contents:
- Assessment of 'epigenetic age' (Horvath et al. 2013, Genome Biol) to see if DNA methylation age differs between matched BTICs vs GBM tumours

### A. Set up wd & packages

We will initially set up our working directory and load our libraries.
```{r include=FALSE}
setwd("/home/sislam/BTIC_GBM")
library(methylumi)
library(gplots)
library(marray)
library(lumi)
library(lattice)
library("RColorBrewer")
library(knitr)
library(xtable)
library(limma)
library(ggplot2)
library(reshape)
```

### B. Load data

We will load the raw 450K data (un-normalized) for the Cancer.dat dataset as well as the clean, filtered, finalized 450K Cancer.dat dataset. Will subset raw 450K data to samples contained in final 450K data and use updated meta data from the final 450K dataset. 

```{r load_data}
load("BTIC_GBM.RData") # load raw 450K methylumi data object for Cancer.dat
dim(betas(BTIC_GBM)) # 485,577 probes, n = 128

load("Cancer.dat.cor.RData") # load final 450K methylumi data object for Cancer.dat
dim(betas(cancer.dat.cor)) # 420,195 probes, n = 70

# subset raw data to just samples contained in the final Cancer.dat object
final.samples<-sampleNames(cancer.dat.cor)
Raw.cancer.dat.sub<-BTIC_GBM[,final.samples]
dim(Raw.cancer.dat.sub) # 485,577 probes, n = 70
# check that sample order for both objects match
identical(sampleNames(Raw.cancer.dat.sub), sampleNames(cancer.dat.cor)) # TRUE
identical(rownames(pData(cancer.dat.cor)), colnames(betas(Raw.cancer.dat.sub))) # TRUE

# pull raw beta matrix
raw.betas.cancer.dat<-betas(Raw.cancer.dat.sub)
identical(colnames(raw.betas.cancer.dat), rownames(pData(cancer.dat.cor))) # TRUE
```

### C. Re-shape beta data put input into epigenetic age calculator

```{r}
# re-format cortex-only dataset
library(sqldf)
#change the setwd filepath to that of the folder with your data. Note the forward slash
setwd("/home/sislam/BTIC_GBM")
betas.cancer.df<-as.data.frame(raw.betas.cancer.dat)
Probe_ID<-rownames(betas.cancer.df) # character vector of length 485,577
dat0<-cbind(Probe_ID,betas.cancer.df)
str(dat0)
datMiniAnnotation=read.csv("datMiniAnnotation.csv")
match1=match(datMiniAnnotation[,1], dat0[,1])
dat0Reduced=dat0[match1,]
head(dat0Reduced)
dat0Reduced[,1]=as.character(dat0Reduced[,1])
dat0Reduced[is.na(match1),1]=as.character(datMiniAnnotation[is.na(match1),1])
datout=data.frame(dat0Reduced)
# make sure you output numeric variables...
for (i in 2:dim(datout)[[2]]  ){datout[,i]= 
as.numeric(as.character(gsub(x=datout[,i],pattern="\"",replacement=""))) }
head(datout)
corrected.datout.colnames<-c("Probe_ID", colnames(betas.cancer.df))
colnames(datout)<-corrected.datout.colnames
str(datout)
# write.table(datout,"Betas.cancer.Horvath.csv", row.names=F, sep="," )
```

### D. Select needed variables for meta data to input into epigenetic age calculator

Use meta data from final 450K Cancer.dat object
```{r}
## for cancer.dat dataset meta data
meta.cancer<-pData(cancer.dat.cor)
colnames(meta.cancer)
head(meta.cancer)
meta.cancer$Sample_Name<-rownames(meta.cancer)
meta.cancer.Horvath<-subset(meta.cancer, select=c("Sample_Name","Age", "Sex"))
# re-format data according to Horvath's specifications in http://labs.genetics.ucla.edu/horvath/dnamage/TUTORIALonlineCalculator.pdf
colnames(meta.cancer.Horvath)<-c("Sample_ID","Age", "Female")
meta.cancer.Horvath$Female<-gsub("f", "1", gsub("m", "0", meta.cancer.Horvath$Female))
meta.cancer.Horvath$Female<-as.numeric(meta.cancer.Horvath$Female)
meta.cancer.Horvath$Tissue<-"Brain FCTX" 
identical(colnames(betas.cancer.df), meta.cancer.Horvath$Sample_ID) #TRUE
meta.cancer.Horvath
# write.table(meta.cancer.Horvath,"Meta.cancer.Horvath.csv", row.names=F, sep=",")
```

## E. Analysis of output from online epigenetic age calculator

```{r}

## load outputs (as csv files)
cancer.output<-read.csv("Cancer.dat.Horvath.Clock.output.csv", header=T) # cancer dataset analysis

cancer.output$SampleID<-meta.cancer.Horvath$Sample_ID # put in corrected sample labels
cancer.output.rev<-cbind(cancer.output, meta.cancer) # append all meta data
colnames(cancer.output.rev)

### analyze cancer-only dataset output

## scatterplot of DNAmAge and Chronological Age (Reported)
ggplot(cancer.output.rev, aes(Age, DNAmAge)) + geom_point(aes(color = AgeAccelerationResidual)) + geom_smooth(method="lm") +
  theme_bw()+xlab("Chronological Age")+ylab("DNA Methylation Age") 

## test for correlation btw DNAm Age and Chronological Age
cor.test(cancer.output.rev$Age, cancer.output.rev$DNAmAge, method= "spearman") # DNAm Age and Age not significantly correlated (Spearman's rho = 0.17, p-value = 0.141)
```

We observe very poor correlation between DNAm age and chronological age with very high DNAm age measures overall. We also see that there appears to be one sample (53_cell) with an abnormally high DNAm age measure of 226. This sample had major warnings from the epigenetic clock output ("Old DNAm age. MAJOR Warning: The methylation levels of this sample have a very low correlation with the expected values: corSampleVSgoldstandard<0.75. Warning: Poorly calibrated methylation levels: meanAbsDifferenceSampleVSgoldstandard >0.15. "). Will remove this outlier BTIC sample and its corresponding tumour sample (53_cell and 53_tumour to be removed). 

```{r}

## remove 53_cell (outlier) and corresponding 53_tumour sample
outlier.samples<-c("53_cell", "53_tumour")
dim(cancer.output.rev.cor<-cancer.output.rev[!(cancer.output.rev$SampleID%in%outlier.samples),])
  
## re-analyze scatterplot of DNAmAge and Chronological Age (Reported)
ggplot(cancer.output.rev.cor, aes(Age, DNAmAge)) + geom_point(aes(color = Sample_Group)) + geom_smooth(method="lm", colour = "gray14") +  scale_color_manual(values=c("#fc8d59","#99d594")) + 
  theme_bw() + xlab("Chronological Age")+ ylab("DNA Methylation Age") 

## test for correlation btw DNAm Age and Chronological Age
cor.test(cancer.output.rev.cor$Age, cancer.output.rev.cor$DNAmAge, method= "spearman") # DNAm Age and Age not significantly correlated (Spearman's rho = 0.19, p-value = 0.1177)

## subset down to cell and tumour subsets to assess DNAm age correlation to chronological age

dim(BTIC.output.cor<-cancer.output.rev.cor[which(cancer.output.rev.cor$Sample_Group=="cell"),])
BTIC.output.cor<-BTIC.output.cor[order(BTIC.output.cor$Patient_ID),] # order by patient ID
dim(Tumour.output.cor<-cancer.output.rev.cor[which(cancer.output.rev.cor$Sample_Group=="tumour"),]) # order by patient ID
Tumour.output.cor<-Tumour.output.cor[order(Tumour.output.cor$Patient_ID),]
identical(BTIC.output.cor$Patient_ID, Tumour.output.cor$Patient_ID) # TRUE

## re-analyze scatterplot of DNAmAge and Chronological Age (Reported) for BTICs
ggplot(BTIC.output.cor, aes(Age, DNAmAge)) + geom_point(aes(color = AgeAccelerationResidual)) + geom_smooth(method="lm") +
  theme_bw()+xlab("Chronological Age")+ylab("DNA Methylation Age")+ ggtitle("Correlation of DNA methylation age and chronological age in BTICs")

## test for correlation btw DNAm Age and Chronological Age for BTICs
cor.test(BTIC.output.cor$Age, BTIC.output.cor$DNAmAge, method= "spearman") # DNAm Age and Age not significantly correlated (Spearman's rho = 0.13, p-value = 0.4514)

## re-analyze scatterplot of DNAmAge and Chronological Age (Reported) for Tumours
ggplot(Tumour.output.cor, aes(Age, DNAmAge)) + geom_point(aes(color = AgeAccelerationResidual)) + geom_smooth(method="lm") +
  theme_bw()+xlab("Chronological Age")+ylab("DNA Methylation Age") + ggtitle("Correlation of DNA methylation age and chronological age in Tumours")

## test for correlation btw DNAm Age and Chronological Age for Tumours
cor.test(Tumour.output.cor$Age, Tumour.output.cor$DNAmAge, method= "spearman") # DNAm Age and Age not significantly correlated (Spearman's rho = 0.22, p-value = 0.2106)
```

After removal of outlier sample (53_cell and her corresponding 53_tumour), we can see that there is poor correlation (Spearman's rho = 0.19, p-value = 0.1177) between DNA methylation age and chronological age. This is consistent with previous work which states " the correlation between DNAm age and chronological age tends to be weak (cor = 0.15, P = 1.9E-29; Additional file 13A)." (Horvath 2013 Genome Biology). The fact that our correlation is not significant is likely to due with the low sample size of our cohort. Below, we will further investigate parameters related to DNA methylation age (ie age acceleration) in these samples.

```{r}
## look at DNAm Age across all matched samples
ggplot(cancer.output.rev.cor, aes(Sample_Group, DNAmAge))+
  geom_boxplot(aes(fill=Sample_Group),color="black", outlier.shape=NA)+  scale_fill_manual(values=c("#fc8d59","#99d594")) +
  geom_point(shape=21, fill="grey",colour="black", size=4, position=position_jitter(w=0.2))+
  theme_bw()+xlab("Origin")+ylab("Predicted DNA Methylation Age")  


wilcox.test(BTIC.output.cor$DNAmAge, Tumour.output.cor$DNAmAge, paired=TRUE) # p-value = 8.127e-05 (Wilcox signed rank test)


## DNAm Age by subtype in BTICs
ggplot(BTIC.output.cor, aes(Parental_Tumour_Subtype, DNAmAge))+
  geom_point(aes(fill = Parental_Tumour_Subtype), shape=21,colour="black", size=4, position=position_jitter(w=0.2))+
  theme_bw()+xlab("Subtype")+ylab("Predicted DNA Methylation Age") 

kruskal.test(DNAmAge ~ Parental_Tumour_Subtype, data = BTIC.output.cor) # p = 0.5848


## DNAm Age by subtype in Tumours
ggplot(Tumour.output.cor, aes(Parental_Tumour_Subtype, DNAmAge))+
  geom_point(aes(fill = Parental_Tumour_Subtype), shape=21,colour="black", size=4, position=position_jitter(w=0.2))+
  theme_bw()+xlab("Subtype")+ylab("Predicted DNA Methylation Age") 

kruskal.test(DNAmAge ~ Parental_Tumour_Subtype, data = Tumour.output.cor) # p = 0.4161
 
```

We can see that BTICs exhibit significantly higher DNA methylation age over matched GBM tumours (p-value = 8.127e-05 (Wilcox signed rank test)). There does not appear to be any association between DNA methylation age and parental tumour subtype. 

```{r}

## look at DNAm Age across all matched samples
ggplot(cancer.output.rev.cor, aes(Sample_Group, AgeAccelerationResidual))+
  geom_boxplot(aes(fill=Sample_Group),color="black", outlier.shape=NA)+  scale_fill_manual(values=c("#fc8d59","#99d594")) +
  geom_point(shape=21, fill="grey",colour="black", size=4, position=position_jitter(w=0.2))+
  theme_bw()+xlab("Origin")+ylab("Age Acceleration Residual")  


wilcox.test(BTIC.output.cor$AgeAccelerationResidual, Tumour.output.cor$AgeAccelerationResidual, paired=TRUE) # p-value = 8.127e-05 (Wilcox signed rank test)


## DNAm Age by subtype in BTICs
ggplot(BTIC.output.cor, aes(Parental_Tumour_Subtype, AgeAccelerationResidual))+
  geom_point(aes(fill = Parental_Tumour_Subtype), shape=21,colour="black", size=4, position=position_jitter(w=0.2))+
  theme_bw()+xlab("Subtype")+ylab("Age Acceleration Residual") 

kruskal.test(AgeAccelerationResidual ~ Parental_Tumour_Subtype, data = BTIC.output.cor) # p = 0.5834


## DNAm Age by subtype in Tumours
ggplot(Tumour.output.cor, aes(Parental_Tumour_Subtype, AgeAccelerationResidual))+
  geom_point(aes(fill = Parental_Tumour_Subtype), shape=21,colour="black", size=4, position=position_jitter(w=0.2))+
  theme_bw()+xlab("Subtype")+ylab("Age Acceleration Residual") 

kruskal.test(AgeAccelerationResidual ~ Parental_Tumour_Subtype, data = Tumour.output.cor) # p = 0.316
```

We can see that BTICs exhibit significantly higher age acceleration over matched GBM tumours (p-value = 8.127e-05 (Wilcox signed rank test)). There does not appear to be any association between age acceleration and parental tumour subtype. 

```{r}
Horvath.sites.df<-read.csv("Horvath_Clock_353_Sites.csv")
Horvath.sites<-as.character(Horvath.sites.df$CpG)

# Subset final cancer.dat 450K data to 353 clock sites in Horvath's epigenetic clock

dim(cancer.clock.dat<-cancer.dat.cor[which(featureNames(cancer.dat.cor)%in%Horvath.sites),]) # 333 probes, n = 70

## Heatmap

cdat <- scale(t(exprs(cancer.clock.dat)))
# define colour palette
BuPu <- colorRampPalette(brewer.pal(10,"RdYlBu")[1:10])(30)
# define coloured variables

tissueCol <- as.numeric(factor(cancer.clock.dat$Tissue))
tissueCol <- gsub("1", "#fc8d59",  gsub("2", "#99d594", tissueCol))
heatmap.2(cdat,
          trace = "none", col = BuPu, dendrogram = "both",
           cexRow = 0.5,
          RowSideColors = tissueCol,
          cexCol = 0.5,
          keysize = 1)
legend("topright",legend = c("cell", "tumour"), 
    col = c("#99d594", "#fc8d59")
)

```

Heatmap of Hovarth clock 333 CpGs (out of 353 CpGs) in the finalized cancer.dat 450K data shows that these CpGs cluster by tissue type. 